---
title: "Mail-in ballots Democrats vs. Republicans"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---


```{r setup, include=FALSE}
# Set up
library(tidyverse)
library(ggplot2)
library(rvest)
library(patchwork)
library(plotly)
library(geepack)
library(viridis)
library(geepack)
library(hrbrthemes)

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)

theme_set(theme_minimal() + theme(legend.position = "bottom"))

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)

scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d

## Comparison of both candidates' mail-in votes

#Compare the number of mail-in ballots of dem/rep in each state

returnedballots_df =
  read_csv("data/mail_in.csv") %>% 
  select(state, party, returned_ballots) %>% 
  filter(party %in% c("Democrats", "Republicans")) %>%
  group_by(state, party) 
 
plot_returned_ballots = 
  returnedballots_df %>%
  ggplot(aes(x = state, y = returned_ballots, fill = party)) +
  geom_bar(stat="identity") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
  title = "The number of returned mail-in ballots of each party by state",
  x = "State",
  y = "Number of returned mail-in ballots",
  caption = "Data from 'https://electproject.github.io/Early-Vote-2020G/'"
  )

#Compare the return rate of mail-in ballots for each party by state
returnrate_df =
  read_csv("data/mail_in.csv") %>% 
  select(state, party, return_rate) %>% 
  filter(party %in% c("Democrats", "Republicans")) %>%
  group_by(state, party) 
 
plot_return_rate = 
  returnrate_df %>%
  ggplot(aes(x =return_rate, y = state, fill = party)) +
  geom_bar(stat="identity", position=position_dodge()) +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +
  labs(
  title = "The return rate of mail-in ballots for each party by state",
  x = "Return rate of mail-in ballots",
  y = "State",
  caption = "Data from 'https://electproject.github.io/Early-Vote-2020G/'"
  )

knitr::opts_chunk$set(
  fig.width = 6,
  fig.asp = .6,
  out.width = "90%"
)
# Create tables of returned ballots and return rates
tab_returnedballots=
    read_csv("data/mail_in.csv") %>% 
  select(state, party, freq_distribution) %>% 
pivot_wider(
    names_from = party,
    values_from = freq_distribution
  ) %>% 
  select(state, Democrats, Republicans) 



tab_returnrate=
    read_csv("data/mail_in.csv") %>% 
  select(state, party, return_rate) %>% 
pivot_wider(
    names_from = party,
    values_from = return_rate
  ) %>% 
  select(state, Democrats, Republicans)

```

# Overview

***

In order to analyze the distribution of mail-in ballots in swing states, we compared the number of mail-in ballots of Democratic Party and the Republican Party in 6 swing states in 2020's election. The 6 swing states are: Florida, Pennsylvania, North Carolina, Iowa, Arizona, and Nevada.

After analyzing the distribution and return rate of mail-in ballots for both candidates in swing states, we can safely draw on our conclusion that Democrats have absolute advantage in 6 swing states in terms of mail-in ballots, given the return rate of mail-in ballots for both candidates are approximately very close to each other.

Relevant outputs of bar charts and tables are attached.

\

# Research question

***

In this section, our goal is to compare the number of mail-in ballots of two candidates' parties in 6 swing states of 2020's U.S. Presidential election. 

We made a frequency table and a bar chart to compare the distribution of mail-in votes of both Democrat and Republican candidates in each state. In addition, we produced another bar chart and a table to compare the return rates of mail in ballots in each state and studied how much more mail-in votes are won by the Democrats in each swing state

\

# Results

***

```{r}
plot_returned_ballots
```

```{r}
knitr::kable(tab_returnedballots)

```

According to plot above, Democrat's candidate Joe Biden, had dominant votes over Republican's candidate Donald Trump, for the returned mail-in ballots in all 6 swing states. The frequency distribution further confirms these facts. In particularly, Biden had absulte dominance over Trump for returned mail-in ballots in the state of Pennsylvania and Florida.

```{r}
plot_return_rate
```

```{r}
knitr::kable(tab_returnrate)
```


The table and bar chart above compared the return rates of both parties' mail-in ballots in 6 swing states.

We observed that return rate of mail-in ballots by parties in each of the swing states are often close to each other, except for the state of Nevada, where Democrats have 41.9% return rate and Republicans have 26.3% return rate. Overall, Democrat's mail-in ballots has higher return rate over Republican's return rate, except for the state of North Carolina, where Republicans have slight advantage of 69.8% while Democrates have 65.5%. 

\

# Discussion

***

Referring back to 2016's election, Democrats obviously gained more support in swing states. Based on the analysis above, we have sufficient evidence to prove that the Democratic party led in all 6 swing states in terms of mail-in ballots. Moreover, return rates of mail-in ballots for both candidates often have small difference. We came to a reasonable interpretation of how Joe Biden became the president-elect of 2020's election in the U.S.

\

# Conclusion

***

After analyzing the distribution and return rate of mail-in ballots for both candidates in swing states, we can safely draw our conclusion on the fact that mail-in ballots in swing states had significantly tilted the 2020 presidential election results. More electorial votes went to Democrats as more mail-in ballots are being counted.

In general, Democrats have absolute advantage in 6 swing states in terms of mail-in ballots, given the return rate of mail-in ballots for both candidates are approximately very close to each other.
