---
title: "Regression Analysis"
output: 
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
---

```{r include = FALSE}
library(tidyverse)
library(httr)
library(patchwork)
library(lubridate)
library(geepack)
library(hrbrthemes)

# setting
theme_set(theme_minimal() + theme(legend.position = "bottom"))

knitr::opts_chunk$set(
  fig.width = 7,
  fig.asp = .7,
  out.width = "70%"
)

options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis",
  digits = 7
)

scale_colour_discrete = scale_color_viridis_d
scale_fill_discrete = scale_fill_viridis_d

# Read data
covid_df = read.csv("./data/covid.csv") 
poll_df = read_csv("./data/state_poll_20.csv") %>% 
  drop_na() 
  

covid_df = covid_df %>% mutate(
  date = as.Date(date)
)


# function that outputs covid df in the interval

match = function(interval, state_name) {
  covid_df %>%
    filter(state == state_name) %>% 
    mutate(logic = date %within% interval) %>% 
    filter(logic == TRUE)
}

# function that generates means of case, death, postive increase, and death increase

mean_value = function(df){
  tibble(
    mean_case = mean(df$total_case),
    mean_death = mean(df$death),
    mean_pos_increase = mean(df$positive_increase),
    mean_death_increase = mean(df$death_increase),
  ) 
}

# Create the regression df

poll_covid_df = 
  poll_df %>% 
  mutate(
    interval = interval(ymd(start_date), ymd(end_date)),
    covid_list = map2(.x = interval, .y = state, ~ match(.x, .y)),
    mean_outputs = map(.x = covid_list, ~ mean_value(.x))
  ) %>% 
  select(state, biden, biden_lead, mean_outputs) %>% 
  unnest(mean_outputs) 

poll_covid_df =
  poll_covid_df %>% 
  mutate(
    id = min_rank(state)
  )

```

\

# Overview

***

In the regression analysis, we found that the supporting rate led by Biden over Trump in the seventeen battlegrounds states reported by all opinion polls slightly decreased as the COVID-19 pandemic progress, suggesting that Trump was more somewhat "benefiting" from COVID-19. This could be due to the **"rally round the flag" effect**, which illustrates the phenomenon that there is a short-term surge of voter approval of present nation leader during a crisis or emergency situation.

The 17 Battleground states include Florida, Pennsylvania, Wisconsin,North Carolina, Michigan, Ohio, Minnesota, Iowa, Arizona, Nevada, Texas, Georgia, Virginia, New Hampshire, Maine, Colorado, and New Mexico. 

Visualization of the change in COVID status and supporting rates over time for each state can be checked [here](https://yikezhao.shinyapps.io/time_trend/).

\

# Research question

***

In this part, we aimed to examine the relationship between COVID-19 case / death rate and presidential poll by different battleground states. Since the COVID-19 statistics was a longitudinal measurement over time, generalized estimating equations (GEE) model was used to account for correlated data within each battleground state. 

\ 

# Distribution of Outcome: Biden's lead over Trump

***

* `biden_lead`: the difference between the supporting rate of Biden and Trump during the poll survey period

The following density plot shows the distribution of `biden_lead`, which is approximately normal.

```{r}

ggplot(data = poll_covid_df, aes(x = biden_lead)) +
  geom_density() + 
  xlab("Biden's lead over Trump (point)")

```

\

# Explanatory variables

***

* `mean_case` the average number of incremental COVID-19 cases between the poll survey period

* `mean_death` the average number of incremental COVID-19 death between the poll survey period

* `mean_pos_increase` the average number of daily COVID-19 cases increased comparing to previous day between the poll survey period

* `mean_death_increase` the average number of daily COVID-19 death increased comparing to previous day between the poll survey period

\

# Result

***

We tested 4 simple GEE models as below:

1. _Biden lead = β0 + β1(COVID case)_

```{r}
# Model mean_case
model_case = 
poll_covid_df %>% 
  drop_na(mean_case, biden_lead)

geeglm(biden_lead ~ mean_case,
       data = model_case,
       id = id,
       family = gaussian(link = "identity"),
       corstr = "exchangeable"
       ) %>%   
  broom::tidy() %>% 
  knitr::kable()

```

Visualization of the predicted and actual lead of Biden against mean COVID-19 case during the polling period:

```{r}
gee_case = geeglm(biden_lead ~ mean_case,
       data = model_case,
       id = id,
       family = gaussian(link = "identity"),
       corstr = "exchangeable"
       ) 

prediction_case = 
  predict(gee_case, type = 'response') %>% 
  as.vector()

model_case = 
  model_case %>% 
  mutate(
    lead_predicted = prediction_case
  )

LineColor <- "#69b3a2"
PointColor <- rgb(0.2, 0.6, 0.9, 1)

ggplot(model_case, aes(x = mean_case)) +
  xlab("Mean number of COVID cases during the polling period") +
  geom_line(aes(y = lead_predicted), color = LineColor) +
  geom_point(aes(y = biden_lead), color = PointColor) +
  scale_y_continuous(
    name = "Predicted lead of Biden over Trump (point)",
    sec.axis = dup_axis(
      name = "Actual lead of Biden over Trump (point)"
      )
  ) +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = LineColor, size = 14),
    axis.title.y.right = element_text(color = PointColor, size = 14),
    axis.title.x.bottom = element_text(size = 14)
  )
```

\

2. _Biden lead = β0 + β1(COVID death)_

```{r}
# Model mean_death
model_death = 
poll_covid_df %>% 
  drop_na(mean_death, biden_lead)

geeglm(biden_lead ~ mean_death,
      data = model_death,
      id = id,
      family = gaussian(link = "identity"),
      corstr = "exchangeable"
      ) %>%   
  broom::tidy() %>% 
  knitr::kable()
```
 
Visualization of the predicted and actual lead of Biden against mean COVID-19 death during the polling period:

```{r}
gee_death = geeglm(biden_lead ~ mean_death,
       data = model_death,
       id = id,
       family = gaussian(link = "identity"),
       corstr = "exchangeable"
       ) 

prediction_death = 
  predict(gee_death, type = 'response') %>% 
  as.vector()

model_death = 
  model_death %>% 
  mutate(
    lead_predicted = prediction_death
  )

ggplot(model_death, aes(x = mean_death)) +
  xlab("Mean number of COVID deaths during the polling period") +
  geom_line(aes(y = lead_predicted), color = LineColor) +
  geom_point(aes(y = biden_lead), color = PointColor) +
  scale_y_continuous(
    name = "Predicted lead of Biden over Trump (point)",
    sec.axis = dup_axis(
      name = "Actual lead of Biden over Trump (point)"
      )
  ) +
  theme_ipsum() +
  theme(
    axis.title.y = element_text(color = LineColor, size = 14),
    axis.title.y.right = element_text(color = PointColor, size = 14),
    axis.title.x.bottom = element_text(size = 14)
  ) 

```

\

3. _Biden lead = β0 + β1(COVID case increase)_

```{r}
# case increase
model_case_increase = 
poll_covid_df %>% 
  drop_na(mean_pos_increase, biden_lead)

geeglm(biden_lead ~ mean_pos_increase,
      data = model_case_increase,
      id = id,
      family = gaussian(link = "identity"),
      corstr = "exchangeable"
      ) %>%   
  broom::tidy() %>% 
  knitr::kable()
```

4. _Biden lead = β0 + β1(COVID death increase)_

```{r}
# death increase
model_death_increase = 
poll_covid_df %>% 
  drop_na(mean_death_increase, biden_lead)

geeglm(biden_lead ~ mean_death_increase,
      data = model_death_increase,
      id = id,
      family = gaussian(link = "identity"),
      corstr = "exchangeable"
      ) %>%   
  broom::tidy() %>% 
  knitr::kable()
```

\

# Model diagnostic

***
**Residual distribution**

In both gee model of COVID-19 case and death, the residuals are approximately normally distributed.

```{r}
case_residual = resid(gee_case, type='response') %>% as.vector()

model_case = model_case %>% mutate(
  residual = case_residual
)

resid_p1 = model_case %>% ggplot(aes(x=residual)) + geom_density()+
    labs( 
      title = "Residual Distribution in Gee Model of COVID-19 Case",
      x = "Residual",
      y = "Density"
      ) + 
    theme(plot.title = element_text(hjust = 0.5))


death_residual = resid(gee_death, type='response') %>% as.vector()

model_death = model_death %>% mutate(
  residual = death_residual
)

resid_p2 = model_death %>% ggplot(aes(x=residual)) + geom_density()+
    labs( 
      title = "Residual Distribution in Gee Model of COVID-19 Death",
      x = "Residual",
      y = "Density"
      ) +
    theme(plot.title = element_text(hjust = 0.5))

resid_p1/resid_p2
```

**Predicted Y against residual**

The residuals of both gee models are evenly distributed around the expected Y. However, we noticed that when the supporting difference is larger, the residuals are more separated, indicating divergence between real supporting difference and predicted model.

```{r message=FALSE}

model_case %>% 
  ggplot(aes(x = lead_predicted,y = residual)) +
  geom_point(color = PointColor) +
  geom_smooth(color = LineColor) +
    labs( 
      title = "Residual vs. Predicted Lead in Case Model",
      x = "Predicted Lead of Biden over Trump",
      y = "Residual"
      ) +
  theme(plot.title = element_text(hjust = 0.5))
  

model_death %>% 
  ggplot(aes(x = lead_predicted,y = residual)) +
  geom_point(color = PointColor) +
  geom_smooth(color = LineColor) +
    labs( 
      title = "Residual vs. Predicted Lead in Death Model",
      x = "Predicted Lead of Biden over Trump",
      y = "Residual"
      ) +
  theme(plot.title = element_text(hjust = 0.5))
        
```


# Findings
***

Based on the regression analysis, we have concluded that:

1. As COVID-19 cases increased, the difference of supporting rate between Biden and Trump decreased within each battleground state. For every 10,000 increase in COVID-19 case, the difference of supporting rate decrease by 0.8 point. The decrease was statistically significant.

2. As COVID-19 death increased, the difference of supporting rate between Biden and Trump decreased within each battle ground state. For every 1,000 increase in COVID-19 death, the difference of supporting rate decrease by 0.3 point. The decrease was statistically significant.

3. COVID-19 daily case / daily death was not correlated with the difference of supporting rate between Biden and Trump.
\

# Discussion

***

Interestingly, Biden supporting rate decreased by increase of COVID-19 cases and death over time. In other word, Trump supporting rate was increasing. This may be due the **"rally round the flag" effect**.

"Rally round the flag" effect is when there is a short-term surge of voter approval of present nation leader during a crisis or emergency situation. One of the most well-known example is during the 9-11 terrorist attack in 2001. The approval rate of President George W. Bush increased by 39% (from 51% to 90%).

According to [Dickinson (2020)](https://sites.middlebury.edu/presidentialpower/2020/03/28/trump-covid-19-and-the-rally-round-the-flag-phenomenon/), the main reason for this effect is the presidents' relatively unique position in the American political system. Presidents is the only individual embodiment of the nation besides the partisan position. Therefore, the president can benefit from his stature as political head of government when circumstances threaten the nation.

He also pointed out that voters who are less politically aware are more responsive to the "rally round the flag" effect, which may be the majority of battleground state voter. Those partisan members are more likely to be ideological.

During the 2020 COVID-19 pandemic, this effect was also observed in several other countries including Australia, Canada, Germany, Britain, India, France, etc. 

\

# Conclusion

***

We found that as the number of COVID-19 cases and deaths increased, Trump's supporting rate slightly increased in the seventeen battleground states.

The GEE regression model we built is relatively simple and we only considered four variables related to COVID status, so this regression analysis should be regarded as an initial and exploratory analysis to generate future hypotheses. Future investigations should consider other variables and potential confounders. 